\propertitle

### 2019-07-10 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-09T16:57:47+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The method introduced in the present work reveals an issue with this
application at the sample level. The issue doesn't come from the method
itself but from a general fact of probability theory. If a probability
distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically it corresponds to the general
impossibility of solving a system of $n$ equations in $m$ unknowns when
$n>m$ \citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In our example this means that if means and pairwise correlations or the
first four moments are informationally sufficient for a particular sample
from a brain area, then they \emph{cannot} be sufficient for the full
network of neurons in that area, or for a different sample. Vice versa, if
we assume that they are informationally sufficient for the full network,
then we must expect them \emph{not} to be sufficient for any recorded
sample thereof. Now, questions about sufficient statistics are meant to be
addressed to a whole brain area, not just to some casually recorded sample.
Thus, paradoxically, maximum-entropy methods applied at the sample level
give an answer opposite to the one we might think they're giving.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
%comparison_ME_sample_full_levels.nb
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. Figure~\ref{fig:comparisons_marginals} gives a graphical answer to
this question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{myblue}{\textbf{------}}). The lower plot
shows the maximum-entropy distributions for the sample, from two moments
(\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{myblue}{$\bm{\square}$}), and the distributions obtained by
marginalizing the full-network distribution to the sample size, from two moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$})
and from four moments (\textcolor{myblue}{$\bm{\blacksquare}$}). The
measured frequency distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also
shown. We note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{myblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.

@@

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-10T10:12:32+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The method introduced in the present work reveals an issue with this
application at the sample level, however. The issue doesn't come from the method
itself but from a general fact of probability theory. If a probability
distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically it corresponds to the general
impossibility of solving a system of $n$ equations in $m$ unknowns when
$n>m$ \citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In our example this means that if means and pairwise correlations or the
first four moments are informationally sufficient for a particular sample
from a brain area, then they \emph{cannot} be sufficient for the full
network of neurons in that area, or for a different sample. Vice versa, if
we assume that they are informationally sufficient for the full network,
then we must expect them \emph{not} to be sufficient for any recorded
sample thereof. Now, questions about sufficient statistics are meant to be
addressed to a whole brain area, not just to some specific, casually
recorded sample. Thus, paradoxically, maximum-entropy methods applied at
the sample level give an answer opposite to the one we might think they're
giving.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
%comparison_ME_sample_full_levels.nb
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. A correct answer can only be given case by case;
\fig~\ref{fig:comparisons_marginals} gives a graphical answer to this
question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{myblue}{\textbf{------}}). The lower plot
shows the maximum-entropy distributions for the sample, from two moments
(\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{myblue}{$\bm{\square}$}), and the distributions obtained by
marginalizing the full-network distribution to the sample size, from two
moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$}) and from four
moments (\textcolor{myblue}{$\bm{\blacksquare}$}). The measured frequency
distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also shown. We
note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{myblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.

@@

*** diff position of mode in marginals says that full distr is extremely
skewed – not surprising since the base space is bounded.
